\textbf{NBA Party}

### 2017-07-05 cardinal_analytx_presentation/nightlife_and_sports.tex
Added citations and clarifying details.

\begin{frame}   \frametitle{What is the problem?}
  \begin{block}{Do pro athletes party to the detriment of their performance?}
    Anecdotally, yes.\footnote{Duffy ('17), Cestone ('17), Duffy ('16), 
      Price ('15), Concepcion ('15), Miller ('15), Campbell ('15), Ball ('12)}

    It's not unreasonable to think that what happens off the court affects outcome, but!
    \begin{itemize}       \item Can we even measure this?
      \item Are there systematic tendencies?     \end{itemize}
  \end{block} \end{frame}

@@

    Knowing the causal relation between partying and next-day performance might inform
    \begin{itemize}       
      \item How coaches set team-rules, 
      \item How travel schedules are determined.
      \item Which players to bench, and when.
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}   \frametitle{Is the problem hard?}
  Yes, in general finding arbitrage opportunities is hard.

  Further, it's not obvious that we will find an effect.

@@

\begin{frame}
  \frametitle{Previous Research - Player shooting average}     Players under-perform proportion of shots made on Sundays.\footnote{Chase, ('15).}
    \begin{itemize}       \item 1,092 of 2,035 NBA players who took $\geq 1$ shot on Sundays under-perform relative to their own season-long average; $p=0.001$.
      \item Effect size is not considered.
      \item Density of support is questionable.\footnote{What if we miss the only shot we took on Sunday, should that be considered?}     \end{itemize}
\end{frame}

\begin{frame}   \frametitle{Previous Research - Team's ability to meet spread}
  Teams under-perform bookmakers expectations on Sunday afternoons when on the road.\footnote{Ezekowitz, ('14).}
    \begin{itemize}       
    \item Only 129 out of 284 games did the away team meet the spread, 
      using 2008-14 data; $p=0.1089$.
    \item Not quite significant at the 90\% confidence level.     
    \end{itemize}
  \end{frame}

\begin{frame}   
  \frametitle{What is our solution?}
  \begin{block}{We can't observe whether players party.}
    Instead, we consider a quasi-experiment. Motivating questions:
    \begin{itemize}       \item Do teams play worse the day after visiting a ``party city''?
      \item How do we even determine whether a city has nightlife?    
      \item What does it mean to ``play worse''? \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{How can we measure unexpected changes in performance?}     \begin{block}{Bookmaker spreads (NBA)}     
    The expected score differential between two teams.

    Conditions on observed performance, and future expectations.

    A team is expected to meet the spread $\frac{1}{2}$ the time.\footnote{Dubner and Levitt, 2006}
  \end{block}
  \begin{block}{Bookmaker money-lines (MLB)}     The odds that a team will beat their opponent.

@@

\begin{frame}   \frametitle{Quasi Experiment}
  \begin{block}{Consider games played back-to-back within 24 hours.}     
    If next day opponent is uncorrelated with visitation to a party city $\leadsto$ identification.

    Weaker, still: if odds of beating the spread uncorrelated \\
    $\leadsto$ identification.\footnote{Under an efficient market, this assumption should hold.}    \end{block}  
\end{frame}

\begin{frame}   \frametitle{Sports: NBA and MLB}
  \begin{block}{We perform the analysis for both NBA and MLB}     If we yield a causal model, we can realize a betting schema.   \end{block} \end{frame}

@@

  We query \href{http://www.basketball-reference.com/leagues/}{basketball-reference.com}
  for a listing of game-days, and then 
  \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{covers.com} for 
  game outcomes alongside a point-spread.

  \begin{itemize}     \item Missing the first couple months of 2010-11 season.
    \item At most two missing games in each of remaining seasons.
    \item Total of 7,829 games in our dataset.   \end{itemize}
\end{frame}

@@

  We query \href{http://www.baseball-reference.com/leagues/}{baseball-reference.com}
  for a listing of game days, and again look to Covers.com for game outcomes and money-lines.

  \begin{itemize}     \item We have on average 2,100 games per season.
    \item (Missing observations due to broken links?)
    \item Total of 12,709 games in our dataset.    \end{itemize} \end{frame}

@@

  All teams are in the US with the exception of 
  \begin{itemize}     \item Toronto Raptors (NBA)
    \item Toronto Blue Jays (MLB)   \end{itemize}

  These two teams are excluded from our analysis.\footnote{We don't expect Toronto to be a huge party city.} \end{frame}

\begin{frame}   \frametitle{Minutes data for NBA}
  For NBA, we also collect minutes data, which describes the number of 
  \begin{itemize}     \item Free-throws
    \item Field goals
    \item Three-pointers   \end{itemize}
  attempted and made by each player in each game. For each season and each team,
  we have nearly complete minutes data; 85\% of our season-teams' have no more than
  3 missing games. \end{frame}

@@

  \begin{itemize}     
    \item Home team effect
    \item Fatigue from previous game
    \item Jet lag   
    \end{itemize}

  The benefit of using spread data is that it controls for salient factors which affect game outcomes.
  The downside is that we don't know exactly what it controls for. \end{frame}

\begin{frame}   \frametitle{Feature Creation}
  We add features which describe
  \begin{itemize}     
    \item Number of days since the last game (i.e. rest-time),
    \item Lagged changes in possession,
    \item Lagged game location, and
    \item Travel Distance.   
  \end{itemize}
  We seek to construct a model which accounts for travel-fatigue and jet-lag.
  For this, we need a little bit more. \end{frame}

@@

  In order to account for jet-lag, we need to be mindful of the direction of travel. We calculate
  \begin{itemize}     \item distance traveled between games,
    \item final direction (bearing) traveling from $a \leadsto b$, and
    \item direction of travel (in east-west degrees) by taking sine.
  \end{itemize}
  Finally, to account for jetlag we interact this east-west measure with travel distance.\footnote{Lallensack ('17)} \end{frame}

@@

\begin{frame}   \frametitle{Discrete Indicator}
  \begin{block}{Anecdotally, LA and NY appear in news articles frequently.}     This motivates a binary indicator for party when either of these two cities have been visited within 24-hours of another game.
    \[
      \texttt{party} \coloneqq \mathbbm 1 \left \{\substack{\textrm{last game played in LA or NY} \\ \textrm{at most 24-hours ago}}\right \}
    \]   \end{block} \end{frame}

@@

\begin{frame}   \frametitle{Placebo Test: what if we let players rest?}
  \centering
  \tiny{
  \begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Meet the Spread (NBA) \\  \hline \\[-1.8ex]   party placebo & 0.072 \\    & (0.082) \\    & \\   lag changes in posession & 0.00001 \\    & (0.002) \\    & \\   logged travel distance & $-$0.005 \\    & (0.026) \\    & \\   east-west travel direction & $-$0.099 \\    & (0.232) \\    & \\   number hours rest time & 0.001$^{*}$ \\    & (0.001) \\    & \\   time of game during day & 0.001 \\    & (0.013) \\    & \\   home team effect & $-$0.009 \\    & (0.043) \\    & \\   logged travel distance * east-west & 0.015 \\    & (0.036) \\    & \\   Constant & $-$0.070 \\    & (0.378) \\    & \\  \hline \\[-1.8ex]  Observations & 9,823 \\  Log Likelihood & $-$6,806.702 \\  Akaike Inf. Crit. & 13,631.400 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}  
} \end{frame}

\begin{frame}   \frametitle{Differences between NBA and MLB}
  The same analysis can be replicated for MLB. Notable differences:
  \begin{itemize}     \item Money-lines are used instead of point-spreads. 
      So instead of predicting likelihood of meeting the spread,
      we predict likelihood of winning conditional on bookmakers odds.
    \item We interact our measure of nightlife with a weekend indicator.
    \item Games are played in series, so we no longer have the problem of an unknown travel schedule, i.e. players stay overnight in the same location within a series.   \end{itemize} \end{frame}

@@

  \begin{itemize}     \item Avoid using black-box feature
    \item Pinpoint more precisely what elements of game-play are affected by nightlife visitation.   \end{itemize}

  Anecdotally, rebounds are considered a ``hustle'' statistic, in the sense that players
  who work harder on the court and expend energy boxing out other players will have higher
  rebound rates. \end{frame}

@@

  Common knowledge suggests that overall defense of a team is strongly influenced by individual player effort.

  Is it the case the hangover effects are more prevalent on defensive aspects of gameplay? \end{frame}

@@

\begin{frame}   \frametitle{Future Research}
  Within the NBA
  \begin{itemize}     \item Are teams more likely to party after a win or a loss?
    \item Do teams who aren't going to make the playoffs party more?
    \item Are there demographic differences which influence propensity to party? E.g. are younger, unmarried players more susceptible to our treatment effect?
    \item We could also examine how shots taken vs. shots scored are affected for star players.
  \end{itemize}
  A preliminary and ad-hoc analysis shows that injury risk increases as a function of party.

  This might inform coaches which players to bench and when.

  Within the MLB, we can look at statistics such as batting averages or general fielding statistics. \end{frame}

\begin{frame}   \frametitle{Future Research (General)}
  \begin{itemize}     \item Possibly prolonged effects; recall our magic 24 hour number.
    \item European soccer leagues.   \end{itemize} \end{frame}

>>>>

\begin{frame}   \frametitle{What is the problem?}
  \begin{block}{Do pro athletes party to the detriment of their performance?}
    Anecdotally, they do party.\footnote{Duffy ('17), Cestone ('17), Duffy ('16), 
      Price ('15), Concepcion ('15), Miller ('15), Campbell ('15), Ball ('12).}

    \vspace{12pt}It's not unreasonable to think that what happens off the court affects game-outcomes, but!
    \begin{itemize}       \item Can we even measure this?
      \item Are there systematic tendencies?     \end{itemize}
  \end{block} \end{frame}

@@

    \vspace{12pt}Even aside from bettors, knowing the causal relation between partying and next-day performance might inform
    \begin{itemize}       
      \item how coaches set team-rules, 
      \item how travel schedules are determined, as well as
      \item which players to bench, and when.
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}   
  \frametitle{Is the problem hard?}

  \begin{block}{Finding causal-mechanisms which significantly determine game-outcomes is difficult.}       It's not obvious whether we will find an effect.

      \vspace{12pt}After all, we can't even observe if players party, let alone when.

      Even if we could, can we exploit this in the marketplace?   \end{block}

@@

\begin{frame}
  \frametitle{Previous Research - NBA player shooting average}     
  NBA Players under-perform proportion of shots made on Sundays.\footnote{Chase, ('15).}
    \begin{itemize}       
    \item 1,092 of 2,035 NBA players who took $\geq 1$ shot on Sundays under-perform 
      relative to their own season-long average; $p=0.001$.
      \item Effect size is not considered.
      \item Density of support is questionable.\footnote{What if we miss the only shot we took on Sunday, is that relevant?}     \end{itemize}
\end{frame}

\begin{frame}   \frametitle{Prev. Research - Likelihood of meeting the spread in NBA}
  Teams under-perform bookmakers expectations on Sunday afternoons when on the road.\footnote{Ezekowitz, ('14).}
\vspace{12pt}
    \begin{itemize}       
    \item Only 129 out of 284 games did the away team meet the spread, 
      using 2006-14 data; $p=0.1089$.
    \item Not quite significant at the 90\% confidence level.     
    \item If we bet \$100 on each Sunday game, we'd only profit \$1,310 over 8 years.\footnote{$(284 - 129) \times 100 + 129 \times (-110) = 1,310$.} Not exciting.
    \end{itemize}
  \end{frame}

\begin{frame}   
  \frametitle{What is our solution?}
  \begin{block}{We can't observe whether players party.}
    Instead, we will consider a quasi-experiment. Motivating questions:
    \begin{itemize}       \item Do teams play worse the day after visiting a ``party city''?
      \item How do we even determine whether a city has nightlife?    
      \item What does it mean to ``play worse''? \end{itemize}
  \end{block}

  \vspace{12pt}We define ``play worse'' relative to bookmaker/public expectations.
\end{frame}

\begin{frame}
  \frametitle{How can we measure unexpected changes in performance?}     \begin{block}{Bookmaker point-spreads (NBA)}     
    The expected score differential between two teams.

    Conditions on every conceivable element of the game:
    \begin{itemize}     \item strengths and weaknesses against particular opponents,
    \item momentum,
    \item injuries, and
    \item even possibly a weather forecast.     \end{itemize}

    A team is expected to meet the spread $\frac{1}{2}$ the time.\footnote{Dubner and Levitt, ('06). Losers pay 10\% fee on top of wagers; a balanced book guarantees 5\% gain for the house-bookmaker.}
  \end{block}
  \begin{block}{Bookmaker money-lines (MLB)}     The odds that a team will beat their opponent.

@@

\begin{frame}   \frametitle{Identification}
    If next day opponent is uncorrelated with visitation to a party city,
    then looking at the \emph{actual outcome} of the game $\leadsto$ identification.

\vspace{15pt}

    Weaker, still: if odds of beating the spread uncorrelated with previous game location
    $\leadsto$ identification.\footnote{Under an efficient market, this assumption should hold.}     \end{frame}

\begin{frame}   \frametitle{Quasi Experiment}
  Consider games played back-to-back within 24 hours.

  \vspace{12pt}
  Examine whether spread is met as a function of nightlife index for lagged game location.

  \vspace{12pt} (Alcohol and loss of sleep are well correlated with nightlife; both have been shown to have negative effects on sports performance.)\footnote{Shirreffs and Maughan ('06), Barnes ('14), Pilcher and Huffcutt ('96).} \end{frame}

\begin{frame}   \frametitle{Sports: NBA and MLB}
  \begin{block}{We perform the analysis for both NBA and MLB}     
    Observe a team traveling to a city with active nightlife.
    \vspace{12pt}
    \begin{itemize}       \item For NBA teams playing another game in the next 24 hours, are they less likely to meet the points-spread?
      \item For MLB teams, after conditioning on the odds of winning, are they less likely to beat their opponent?     \end{itemize}
    \vspace{12pt}\begin{block}{If we yield a causal model, we may realize a betting schema.}     \end{block}
  \end{block} 
\end{frame}

@@

  \vspace{12pt}We query \href{http://www.basketball-reference.com/leagues/}{basketball-reference.com}
  for a listing of game-days, and then 
  \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{covers.com} for 
  game-outcomes alongside a point-spread.

  \vspace{12pt}  
  \begin{itemize}     \item Missing the first couple months of 2010-11 season.
    \item At most two missing games in each of remaining seasons.
    \item Total of 7,829 games in our dataset.   \end{itemize}
\end{frame}

@@

  \vspace{12pt}We query \href{http://www.baseball-reference.com/leagues/}{baseball-reference.com}
  for a listing of game days, and again look to covers.com for game-outcomes and money-lines.

  \vspace{12pt}
  \begin{itemize}     \item We have on average 2,100 games per season.
    \item (Missing observations due to broken links, postponed games?)
    \item Total of 12,709 games in our dataset.    \end{itemize} \end{frame}

@@

  \vspace{12pt}  All teams are in the US with the exception of 
  \vspace{12pt}\begin{itemize}     \item Toronto Raptors (NBA)
    \item Toronto Blue Jays (MLB)   \end{itemize}

  \vspace{12pt}These two teams are excluded from our analysis.\footnote{We don't expect Toronto to be a huge party city.} \end{frame}

\begin{frame}   \frametitle{Minutes data for NBA}
  For NBA, we also collect minutes data, which describes the number of 
  \vspace{12pt}
  \begin{itemize}     
    \item Free-throws
    \item Field goals
    \item Three-pointers   
  \end{itemize}
  \vspace{12pt}attempted and made by each player in each game. For each season and each team,
  we have nearly complete minutes data; 85\% of our season-teams have no more than 3 missing games. \end{frame}

@@

  \vspace{12pt}\begin{itemize}     
    \item Home team effect
    \item Fatigue from previous game
    \item Jet lag   
    \end{itemize}

  \vspace{12pt}The benefit of using spread data is that it controls for salient factors which affect game-outcomes.
  The downside is that we don't know exactly what it controls for. \end{frame}

\begin{frame}   \frametitle{Feature Creation}
  We add features which describe
  \vspace{12pt}  
  \begin{itemize}     
    \item Number of days since the last game (i.e. rest-time),
    \item Lagged changes in possession,\footnote{A crude measure of fatigue incurred from last game.}
    \item Lagged game location, and
    \item Travel Distance.   
  \end{itemize}
  \vspace{12pt}We seek to construct a model which accounts for travel-fatigue and jet-lag.
  For this, we need a little bit more. \end{frame}

@@

  \vspace{12pt}In order to account for jet-lag, we need to be mindful of the direction of travel. We calculate
  \vspace{12pt}\begin{itemize}     \item distance traveled between games,
    \item final direction (bearing) traveling from $a \leadsto b$, and
    \item direction of travel (in east-west degrees) by taking sine.
  \end{itemize}
  \vspace{12pt}Finally, to account for jetlag we interact this east-west measure with travel distance.\footnote{Lallensack ('17)} \end{frame}

@@

\begin{frame}   \frametitle{Discrete Indicator}
  \begin{block}{Anecdotally, LA and NY appear in news articles frequently.}     \vspace{12pt}This motivates a binary indicator for party when either of these two cities have been visited within 24-hours of another game.
\vspace{12pt}
    \[
      \texttt{party} \coloneqq \mathbbm 1 \left \{\substack{\textrm{last game played in LA or NY} \\ \textrm{at most 24-hours ago}}\right \}
    \]   \end{block} \end{frame}

@@

\begin{frame}   \frametitle{Placebo Test (discrete indicator): what if players rest?}
  \centering
  \tiny{
\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & Meet the Spread (NBA) \\  \hline \\[-1.8ex]   Party placebo & 0.072 \\    & (0.082) \\    & \\   Lag changes in posession & 0.00001 \\    & (0.002) \\    & \\   Logged travel distance & $-$0.005 \\    & (0.026) \\    & \\   East-west travel direction & $-$0.099 \\    & (0.231) \\    & \\   Number hours rest time & 0.001$^{*}$ \\    & (0.001) \\    & \\   Time of game during day & 0.001 \\    & (0.013) \\    & \\   Home team effect & $-$0.009 \\    & (0.043) \\    & \\   Logged travel distance * east-west & 0.015 \\    & (0.036) \\    & \\   Constant & $-$0.069 \\    & (0.378) \\    & \\  \hline \\[-1.8ex]  Observations & 9,823 \\  Log Likelihood & $-$6,806.701 \\  Akaike Inf. Crit. & 13,631.400 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}  } \end{frame}

\begin{frame}   \frametitle{Differences between NBA and MLB}
  The same analysis can be replicated for MLB. Notable differences:
  \vspace{12pt}  
  \begin{itemize}     
    \item Money-lines are used instead of point-spreads. 
      So instead of predicting likelihood of meeting the spread,
      we predict likelihood of winning conditional on bookmakers odds.
    \item We interact our measure of nightlife with a weekend indicator.
    \item Games are played in series, so we no longer have the problem of an unknown travel schedule, i.e. players stay overnight in the same location within a series.   \end{itemize} \end{frame}

@@

  \vspace{12pt}
  \begin{itemize}     
    \item Avoid using black-box feature
    \item Pinpoint more precisely what elements of game-play are affected by nightlife visitation.   
\end{itemize}

\vspace{12pt}  Anecdotally, rebounds are considered a ``hustle'' statistic, in the sense that players
  who work harder on the court and expend energy boxing out other players will have higher
  rebound rates. \end{frame}

@@

  \vspace{12pt}Common knowledge suggests that overall defense of a team is strongly influenced by individual player effort.

  \vspace{12pt}Is it the case the hangover effects are more prevalent on defensive aspects of gameplay? \end{frame}

@@

\begin{frame}   \frametitle{Future Research - NBA}
  Within the NBA
  \begin{itemize}     
    \item Are teams more likely to party after a win or a       loss?
    \item Do teams who aren't going to make the playoffs party more?
    \item Are there demographic differences which influence propensity to party? E.g. are younger, unmarried players more susceptible to our treatment effect?
    \item We could also examine how shots taken vs. shots scored are affected for star players.
  \end{itemize}
  \vspace{12pt}  A preliminary and ad-hoc analysis shows that injury risk increases as a function of party.\footnote{We look for instances where players are put on the bench and stop playing (at least several) games as a function of our nightlife indicator.}
  This might inform coaches which players to bench and when.

\begin{frame}   \frametitle{Future Research (MLB and General)}
  \begin{itemize}     
  \item Within the MLB, we can look at statistics such as batting averages or general fielding statistics. 
    \item Possibly prolonged effects; recall our magic 24 hour number.
    \item European soccer leagues.   \end{itemize} \end{frame}
